Strict Pure Functional Programming and Automatic Parallelization\\ (New Project-in-progess Paper)

### 2016-02-23 pubs/tfp16/paper.tex
syncing on this fine closure day

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (Project-in-progess Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and 
        Dean Prichard\inst{1} \and 
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and 
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\emph{Preliminary paper note: We are reporting work in progress.  At the time
  of writing we have a complete serial implementation and are currently
  developing the parallel runtime.  We expect to have preliminary results from
  our parallel implementation by the time of the symposium.}

@@

\section{Background and Motivation}
The practice of highest-performance scientific computing (HPC) is, overall,
highly conservative with respect to change.  Even as the US government commits
to a push for exascale computing ($10^{18}$ FLOPS) within the next decade, a
highly respected member of the HPC community recently stated ``Fortran is
essential for exascale programming.''~\cite{Heroux16}.  Existing HPC codes run
to hundreds of thousands of lines and complete rewrites are infeasible.  C++
has become more common for new code starts, but their nature remains the same:
huge codes with poor (or non-existent) encapsulation of side effects, and
multiple levels of parallelism also poorly abstracted, especially at the
thread (shared-memory) level.  Thread-level parallel efficiency can be poor
and is explained by Amdahl's law because the parallel model is bulk synchronous.

@@

@@

@@

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (New Project-in-progess Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and 
        Dean Prichard\inst{1} \and 
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and 
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\emph{Preliminary paper note 1: We are reporting work in progress.  At the time
  of writing we have a complete serial implementation, a primitive
  proof-of-concept parallel implementation, and are currently developing a
  more realistic parallel runtime.  We expect to have preliminary results from
  our new parallel implementation by the time of the symposium.}

\emph{Note 2:  While this does not qualify as a student paper, students (undergraduate!)
have contributed considerably to this effort, and if accepted we plan to have one of these students
present this paper.}

@@

\section{Background and Motivation}
The practice of highest-performance scientific computing (HPC) is, overall,
highly conservative with respect to change.  Even as the US government commits
to a push for exascale computing ($10^{18}$ FLOPS) within the next decade, a
highly respected member of the HPC community recently stated ``Fortran is
essential for exascale programming.''~\cite{Heroux16} Existing US government
laboratory HPC codes run to hundreds of thousands of lines and complete
rewrites are infeasible.\footnote{A European consortium is anticipating
  exascale computing as well~\cite{EUexascale}, but the authors are not
  knowledgeable about the state of large-scale, non-USA HPC codes.}  C++ has
become more common for new code starts, but their nature remains the same:
huge codes with fair to non-existent encapsulation of side effects, and
multiple levels of parallelism also poorly abstracted, especially at the
thread (shared-memory) level.  Thread-level parallel efficiency can be poor
and is explained by Amdahl's law because the parallel model is bulk
synchronous.

@@

\subsubsection{Proper tail calls}

Proper tail calling is achieved.  Because we maintain our own data and control
(continuation) stack all generated C functions take no arguments and have
return type void, and with a single exception in the non-strict case are
invoked in tail call positions, both gcc and Clang/LLVM generate jumps instead
of calls for C calls in tail call positions.

The exceptions to tail calling are for convenience.  Similar to what is
described in the \emph{fast curry} paper we have a generic variadic stgApply
function that is used for applications of non-known functions, and known
functions with a deficit or excess number of arguments.  The stgApply function
takes as arguments the first element of the application (FUN, PAP, THUNK, or
BLACKHOLE) and its arguments, and if a THUNK, evaluates it with a call (rather
than jump) to yield a FUN or PAP.  This allows us to implement a single
stgApply function as a single C function.  The implementation of stgApply as
described in \emph{fast curry} makes the same concession.

For stgApply in the non-strict semantics the alternative to call-with-return
would be a pair of functions stgApply1 and stgApply2.  In one possible variation, 
stgApply1 would push a continuation with stgApply2 as the return address, then
invoke the first element of the application.  Call-with-return becomes more
convenient when considering the various forms of strictness:  when evaluating
function or constructor arguments it is simpler, and arguably more efficient
in terms of time (but not space in general) to ``evaluate in place'' than
implement the continuation style to the logical limit.

@@

\section{Acknowledgments}

DOE/NNSA/LANL LDRD, DOE/NNSA SULI.

@@

\bititem{EUexascale}
Toward Exascale Computing.
European Exascale Software Initiative.
http://www.eesi-project.eu/
